\textbf{ TESSEROIDES CON DENSIDADES VARIABLES: CÁLCULO DE CAMPOS GRAVITATORIOS EN COORDENADAS ESFÉRICAS CON DENSIDADES VARIABLES EN PROFUNDIDAD }

### 2017-09-27 jovenes-investigadores/jovenes-investigadores.tex
Started writing breef work for jovenes investigadores

\documentclass[extra]{gji}
%~ \documentclass[extra, referee]{gji}

\usepackage[utf8]{inputenc}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{graphicx}
%~ \usepackage{todonotes} % to make annotations on margins
\usepackage[disable]{todonotes} % to disable annotations on margins

\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
    allcolors=blue,
}

@@

\title[Variable Density Tesseroids]{
    Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de San Juan, Argentina\\
    $^3$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil
    }

@@

\begin{summary}
Summary of this paper 
\end{summary}

\begin{keywords}
Satellite gravity??, Numerical modelling, Structure of the Earth, Gravity anomalies and Earth structure, Numerical approximations and analysis, spatial analysis??
\end{keywords}

\section{Introduction}

Lithosphere's density variation with depth has been studied for almost a century. 
On the first years \citet{Athy1930} obtained a decreasing exponential relation between both quantities by studying shale samples, while in the following decades other density functions have been proposed for different rock types \citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, the density variation with depth has been taken into account in forward and inversion gravity models, mostly applied to grabens and basins \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
A spherical prism, also known as  Tesseroid, in a geocentric spherical coordinate system, with a computation point $P$ and its local north oriented Cartesian coordinate system. Picture made by \citet{Uieda2015}.
}
\label{fig:tesseroid-uieda}
\end{figure}

\citet{Uieda2016} developed a forward gravity model based on tesseroids with homogeneous densities using the GLQ approximation method.
It has been originally implemented through command line programs written in C programming language.
However, it has been ultimately ported to Python and included in the open-source library Fatiando a Terra \citep{Uieda2013} for geophysical modelling and inversion.

In order to develop a forward model that computes the gravity field generated by any tesseroid with arbitrary variable density, the Taylor series expansion is not well suited, because it would need to derive the series expansion terms for each density function.
On the other hand, the GLQ allows us to calculate them for any density function without any other information but the function itself.

\citet{Ku1977} documented that the GLQ becomes less accurate when the computation point is closer to the tesseroid or for smaller GLQ order. 
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it ensures the accuracy implementing a modified version of the adaptive discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain distance-size ratio is lower than a predefined value ($D$) that controls the accuracy of the computation. 
\citet{Uieda2016} have also objectively obtained standard values of $D$ for gravity potential, gradients and tensor components comparing the fields generated by a spherical shell, which constitutes a special case of the volume integrals that has analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}\todo{citar mas??}.

We have enhanced the later implementation with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It's based on the GLQ approximation, makes use of the modified distance-size discretization algorithm and is written in Python and Cython languages, obtaining an easy to use library that runs precompiled code for the more time consuming functions.

In order to ensure the accuracy of the numerical approximation, we have conducted similar tests as the one performed by \citet{Uieda2016} by comparing the results of the numerical model with the analytical solutions for a spherical shell with linear, exponential and discontinuous density.
Furthermore we had to derive the analytical expressions for these density functions.

This new forward gravity model makes use of some previously existing classes and functions from Fatiando a Terra, what saved us time and allowed us to focus on the new enhancements.
Furthermore, we have the intention to include it into a future release of the Fatiando a Terra library in order to make its distribution and maintenance easier.

In the following sections we will describe how the new algorithm works, its theoretical background and obtain the distance-size ratio values needed to get a good accuracy in the computation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Theory}

The spherical prisms known as tesseroids are mass elements defined in a geocentric spherical coordinate system bounded by a pair of parallels, a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the gravity fields generated by the tesseroid are going to be calculated with respect to the local north oriented Cartesian coordinate system located at $P$.
In the special case of homogeneous density, the gravity potential, its gradient and the Marussi tensor components can be calculated through the integrals obtained by \citet{Grombein2013} \citep[see also][]{Uieda2016}.

In our case, we will assume that the tesseroid has an arbitrary variable density in depth, i.e. only depends on the radius spherical coordinate.
Thus, the integrals for the gravity fields are slightly modified:

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}

\noindent where

\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesseroid-tensor-kernel}
\end{equation}

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and

\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

According to \citet[p.~390]{Hildebrand1987}, we can approximate any integral in the interval $[-1, 1]$ using a $N$th order GLQ, i.e. by a weighted sum of the integration kernel evaluated on the roots of the $N$th Legendre polynomial:

\begin{equation}
    \int\limits_{-1}^1 f(x) dx \approx \frac{b-a}{2} \sum_{i=1}^N W_i f(x_i),
\end{equation}

\noindent where $x_i$ are the roots of the $N$th Legendre polynomial $P_N(x)$ and $W_i$ are the weights of each term:

\begin{equation}
    W_i = \frac{2}{(1-x_i^2)[P_N^\prime(x_i)]^2},
\end{equation}

\begin{equation}
    P_N(x_i) = 0, \quad \forall i = {1,\dots,N}.
\end{equation}

In case of an integration defined on a arbitrary interval, e.g. $[a,b]$, we can scale the nodes and perform a similar approximation:

\begin{equation}
    \int\limits_a^b f(x) dx \approx \frac{b-a}{2} \sum_{i=1}^N W_i f(x_i^s),
\label{eq:glq-scaled}
\end{equation}

\noindent where $x_i^s$ is the scaled $x_i$ root in the $[a,b]$ interval, also called nodes of the quadrature:

\begin{equation}
    x_i^s = \frac{b-a}{2} x_i + \frac{b+a}{2}.
\end{equation}

For example, if we use a second order GLQ, the roots of the $P_2(x)$ Legendre polynomial and its corresponding weights are $x_i = \pm 0.577350269$ and $W_i = 1$.

Our intention is to use GLQ to approximate the volume integrals from equations \ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor}. 
When tesseroids have variable densities, we can write the integral kernels as the product of a certain function $g$ and the density function:

\begin{equation}
    f(r', \phi', \lambda') = \rho(r') g(r', \phi', \lambda'),
\end{equation}

\noindent and then apply the quadrature to every integral corresponding to each spherical coordinate, obtaining a similar expression to the one shown by \citet{Asgharzadeh2007} and \citet{Uieda2016}:

\iftwocol{
\begin{equation}
    \begin{split}
        \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') 
        d\Omega \approx& \\
        A 
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        & W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

\noindent where

\begin{equation}
    A = 
    \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8}.
\end{equation}

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, the information of the density function is summarised as the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with a well fitted discretization algorithm, approximates the gravity fields with good accuracy and precision.\todo{no se si poner esta ultima oracion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Analytical Solutions for Spherical Shell}

The integrals that define the gravity fields in equations \ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor} have analytical solutions for the exceptional case of a spherical shell with homogeneous density \citep{Mikuska2006,Grombein2013}.
In order to test the accuracy of their approximations, \citet{Uieda2016} and \citet{Grombein2013} compared their numerical models (tesseroids with homogeneous density) with these analytical solutions.

In case of variable density tesseroids, the comparison must obviously be made with a variable density spherical shell, whose analytical solution will depend on the chosen density function.
Because the literature has not offered it yet, we had to develop it ourselves. \todo{ no se si ponerlo asi}

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, whose density depends on the radius spherical coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$, located along the $z$ axe at a distance $r$ from the origin, can be written as follows:

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
The computation point $Q$ is located in the $z$ axe at a distance $r$ from the centre of the shell.
For practical purposes we will assume that $Q$ is outside the outer radii, i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

\begin{equation}
    V_\text{sh}(r) = G 
    \int\limits_0^{2\pi}
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r')}{\ell} {r'}^2 \cos\phi' \, 
    dr' d\phi' d\lambda',
\end{equation}

\noindent where $\ell$ is defined in equation \ref{eq:ell}.
The computation point $Q$ is located at latitude $\phi=90^\circ$, so $\ell$ and $\cos\psi$ (defined in equation \ref{eq:cospsi}) get simplified:

\begin{equation}
    \cos\psi = \sin\phi', \quad
    \ell = \sqrt{r'^2 - 2 r r' \sin\phi' + r^2}.
\end{equation}

Due to the rotational symmetry along the $z$ axe, the integration in $\lambda'$ is straightforward:

\begin{equation}
    V_\text{sh}(r) = 2\pi G 
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r') {r'}^2 \cos\phi'}{\sqrt{r'^2 - 2 r r' \sin\phi' + r^2}}
    \, dr' d\phi',
\end{equation}

\noindent and the integration in $\phi'$ can be performed independently of the density function.
Using \citet{sagemath} we obtain the following expression:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}(r) = 2\pi G
        \int\limits_{R_1}^{R_2}
        \Big[ & \sqrt{r^2 + r'^2 + 2rr'} - \\
        & \sqrt{r^2 + r'^2 - 2rr'}
        \Big] \frac{r'\rho(r')}{r} \, dr'.
    \end{split}
\label{eq:shell-pot-sqrts}
\end{equation}
}{
\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'}  -
    \sqrt{r^2 + r'^2 - 2rr'} 
    \Big] \frac{r'\rho(r')}{r} \, dr'.
\label{eq:shell-pot-sqrts}
\end{equation}
}

For our purposes we can assume that the computation point $Q$ is outside the outer radii, i.e. $r>R_2$ ($R_2 \geq r'$), and simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

\begin{equation}
    \sqrt{r^2 + r'^2 + 2rr'} = |r + r'| = r + r',
\end{equation}
\begin{equation}
    \sqrt{r^2 + r'^2 - 2rr'} = |r - r'| = r - r',
\end{equation}

\noindent what leads to the following expression of the potential of the spherical shell:

\begin{equation}
    V_\text{sh}(r) = \frac{4\pi G}{r}
    \int\limits_{R_1}^{R_2} {r'}^2 \rho(r') \, dr'.
\label{eq:shell-pot}
\end{equation}

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity field generated by a spherical shell with a variable density in depth on any outside point located in the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

The gradient and the Marussi tensor derived from potentials that depends solely on $r$ has only a few non zero components: the vertical component of the gradient ($g_z$) and the diagonal components of the tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
They can be easily obtained as \citet{Grombein2013} does, even for any density function $\rho(r')$ of our choice:

\begin{equation}
    g_z = \frac{V_\text{sh}(r)}{r},
\end{equation}
\begin{equation}
    g_{xx} = g_{yy} = -\frac{V_\text{sh}(r)}{r^2}, \quad
    g_{zz} = \frac{2V_\text{sh}(r)}{r^2}.
\end{equation}

For practical purposes we are going to obtain the gravity potential for three different functions from the perspective of their variation in $r'$: a linear and an exponential density function.

\subsubsection{Linear Density}

If the density of the spherical shell is a linear function on the radius coordinate, i.e.:

\begin{equation}
    \rho(r') = ar' + b,
\end{equation}

\noindent then the gravity potential generated on any computation point at distance $r$ from its centre can be obtained from the following expression:

\begin{equation}
    V_\text{sh}^\text{lin}(r) = \pi G \left[ 
    a \frac{R_2^4 - R_1^4}{r} +
    b \,\frac{4}{3} \frac{R_2^3 - R_1^3}{r} \right].
    \label{eq:shell-pot-linear}
\end{equation}

The second term on equation \ref{eq:shell-pot-linear} reproduces the potential generated by a spherical shell with homogeneous density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
So it can be read as the combination of a spherical shell with the later homogeneous density and the same shell with variable density $\rho(r') = ar'$.

\subsubsection{Exponential Density}

If, on the other hand, the spherical shell has an exponential density in the radius coordinate,

\begin{equation}
    \rho(r') = Ae^{-(r' - \Delta h)/b},
\end{equation}

\noindent the potential can be written as follows:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab \, e^\frac{\Delta h}{b}
        \Big[
        & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
        & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
        \Big].
    \end{split}
\label{eq:shell-pot-exp}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^\frac{\Delta h}{b}
    \Big[
    & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
    & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
    \Big].
\label{eq:shell-pot-exp}
\end{equation}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Implementation}

We have developed a Python library that computes the gravity fields generated by any Tesseroid with variable density in depth through the GLQ to numerically approximate the volume integrals.
It's freely available under the BSD 3-clause open-source library and can be downloaded from the following repository: \todo{agregar repo, vale la pena poner el link?}

Although it's a Python library, the core functions of the computation are written in Cython, a superset of Python language that allows to generate efficient C code through the Cython compiler.
This produces a precompiled library whose functions can be called from regular Python code, what allowed us to make it compatible with the preexisting open-source library Fatiando a Terra \citep{Uieda2013}.

Its usage is very similar to the code for homogeneous tesseroid already in Fatiando a Terra (version 0.5):

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Define a single tesseroid (or a tesseroid mesh) with the Fatiando a Terra classes.
    \item Add a predefined Python density function to the tesseroid (or each tesseroid in the mesh).
    \item Calculate the desired gravity field (potential, gradient or tensor component) in any set of computation points with our new functions.
\end{enumerate}

The existing code in Fatiando a Terra (v.~0.5) for homogeneous tesseroid is optimised through the Numba compiler.
Due to difficulties when passing a Python function as argument of precompiled Numba methods, we have decided to rewrite them in Cython language, keeping the homogeneous tesseroid calculation and adding the new variable density code.

From equation \ref{eq:glq-var-dens} we saw that the GLQ approximates the volume integrals as the gravitational effect of $N_\lambda N_\phi N_r$ point masses, where $N_i$ are the orders of the quadrature for the integral on each coordinate, with $i \in \{ \lambda, \phi, r \}$.
If we set the GLQ orders to fixed values, it's easy to conclude that the nearer the computation point is to the tesseroid, the accuracy of the approximation is reduced and the point masses effect becomes more noticeable than the tesseroid we want to approximate.
A more complete analysis of this conclusion can be found in \citet{Ku1977, Li2011, Uieda2016}.

Instead of increasing the GLQ orders to improve the accuracy, what leads to a more time consuming computation, adaptive discretization algorithms have been applied in previous works \citep{Li2011, Uieda2016}.
They essentially consist in dividing the tesseroid based on a ratio between the distance from its geometric centre to the computation point and its dimensions.

In order to make the computation more efficient, we use only second order GLQ for each integration while we ensure the accuracy of the method through the application of the modified adaptive discretization algorithm developed by \citet{Uieda2016}.
The modifications made by \citet{Uieda2016} to the original adaptive discretization method \citep{Li2011} can be summed up in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.

More specifically, \citet{Uieda2016} redefined the dimensions of an arbitrary tesseroid: $L_\lambda$, $L_\phi$ and $L_r$. The former ones are the arc-distances measured along its top surface while $L_r$ is the difference between the radii of the top and bottom surfaces.

The adaptive discretization algorithm starts by checking that the following inequality is satisfied:

\begin{equation}
    \frac{d}{L_i} \geq D,
\label{eq:distance-size-ratio}
\end{equation}

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for any spherical coordinate, then the tesseroid must be divided by half in that direction.
A more complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied, the algorithm computes the desired gravity field using the GLQ approximation.

In summary, the distance-size ratio $D$ determines indirectly how many times the tesseroids will be divided, and therefore regulates both the accuracy of the algorithm and its computation time.
In this way the modified version of the adaptive discretization algorithm efficiently improves its accuracy and gives the user more control over the computation: it allows to stop the division of the tesseroids when their dimensions are lower than a certain value (e.g. less than 10 cm), in which case it does not modifies significantly the desired gravitational field.

On the other hand, the value assigned to the distance-size ratio $D$ cannot be easily related to the error of the approximation, thus the choice of the value of $D$ to get an acceptable accuracy is a matter of study.

In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.
We have performed similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Determination of distance-size ratio}

In order to test the new variable density tesseroid code and estimate the error of the approximation, we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

This error estimations are also useful in the determination of the minimum value for the distance-size ratio $D$ such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

We approximated the spherical shell by a mesh of $30^\circ \times 30^\circ$ tesseroids and calculated the gravity potential, the vertical component of the gradient ($g_z$) and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the pole, (2) another one on the Equator, (3) one located at satellite height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found in the Table \ref{tab:grids}.

We computed the three gravity fields  on every point of these four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5.
Then we calculated the maximum difference between these results and the value of the analytical solution at the same height.

Because the gravity fields generated by a tesseroid are approximated by the effect of point masses, the numerical results may vary between computation points at same height but at different latitude or longitude.
That's why we choose for comparison the maximum difference between the results on every point of the grid and the analytical solution.

Finally, we set the acceptable value of $D$ if the corresponding error of the numerical approximation is lower than 0.1\%.

\begin{table}
\caption{
    Description of the synthetic grids on which the comparison of the numerical model against the analytical solutions for a spherical shell was done. This collection of grids tries to include every possible scenario on which the numerical model can present a different behaviour depending on the size of the grid, its location or its height over the Earth surface. Every grid consists in a set of 10$\times$10 points.
}
\label{tab:grids}
\begin{tabular}{lcccc}
    Grid & Size & Lat. extension & Lon. extension & Height \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Equator & $1^\circ \times 1^\circ$ & $0^\circ - 1^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Satellite & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 260km \\
    Big Grid & $30^\circ \times 30^\circ$ & $60^\circ - 90^\circ$ & $0^\circ - 30^\circ$ & 2km \\
\end{tabular}
\end{table}

\subsection{Linear Density}

In first place, we proposed a spherical shell with linear density in such way that it has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

\begin{equation}
    \rho(r') = ar' + c,
    \label{eq:density-linear}
\end{equation}
\noindent with 
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{35000\text{m}},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} - 
        2670\text{kg/m$^3$}}{35000\text{m}} R + 
        2670\text{kg/m$^3$},
\end{equation}

\noindent where $R$ = 6378.137 km is the mean Earth radius.

Then we computed the difference between the numerical model with its corresponding analytical solution (equation \ref{eq:density-linear}) for a thick and a thin spherical shell of 1km and 35km of thickness, respectively.

In Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick} we can see the results for the potential, the vertical component of the gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor.
Because the other two diagonal components of the tensor show similar results as the ones obtained for the $g_{zz}$, we exclude them from the figure, although they can be found in the repository. \todo{chequear}

For the linear density case, we can observe that the calculation of the potential and the $g_z$ component of its gradient needs a value of $D=1$ and $D=2$, respectively, in order to ensure an accuracy of 0.1\%.
On the other hand, the $g_{zz}$ computations show a noticeable difference between the thin and the thick shell: the first one needs a value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of computing the Marussi tensor components in case of a linear density tesseroid.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dlinear-thin-differences.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model and the analytical solution for a thin spherical shell of 1km of thickness with the linear density defined on equation \ref{eq:density-linear}. The computations were performed on the four grids described in Table \ref{tab:grids} and for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-linear-thin}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dlinear-thick-differences.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model and the analytical solution for a thick spherical shell of 35km of thickness with the linear density defined on equation \ref{eq:density-linear}. The computations were performed on the four grids described in Table \ref{tab:grids} and for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-linear-thick}
\end{figure}

\subsection{Exponential Density}

To test the model in case of an exponential density we proposed the following function:

\begin{equation}
    \rho(r') = A e^{-(r' - R)/b}
\label{eq:density-exp}
\end{equation}

\noindent where $A$ = 1kg/m$^3$ and $b$ = 1000km.
We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, whose results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

In the same way we analysed the linear density results, we can summarise that the thin spherical shell presents the more conservative values of the distance-size ratio: $D=1$ for the gravity potential, $D=2$ for its gradient components and $D=8$ for the Marussi tensor components.

Surprisingly both the linear and the exponential density tesseroid models needs the same values of distance-size ratio $D$ in order to assure a good accuracy.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dexp-shifted-thin-differences.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model and the analytical solution for a thin spherical shell of 1km of thickness with the exponential density defined on equation \ref{eq:density-exp}. The computations were performed on the four grids described in Table \ref{tab:grids} and for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-thin}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dexp-shifted-thick-differences.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model and the analytical solution for a thick spherical shell of 35km of thickness with the exponential density defined on equation \ref{eq:density-exp}. The computations were performed on the four grids described in Table \ref{tab:grids} and for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-thick}
\end{figure}

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$.
In order to do this we proposed a set of density functions, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900 kg/m$^{3}$ at the bottom of the spherical shell.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the respective analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

This tests were performed using both the thin and the thick spherical shell.
On Figures \ref{fig:D-exp-power-thin}a and \ref{fig:D-exp-power-thick}a we can see the density functions used to perform these tests for each spherical shell model, respectively, and their results can be seen in \ref{fig:D-exp-power-thin}b and \ref{fig:D-exp-power-thick}b.
Taking into account that lower values of $b$ implies a more varying function, we can see that for extremely varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\% in the thin shell case, and the same happens with $b$=10km in the thick shell case.

>>>>

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[spanish]{babel}
\usepackage[hmargin=2.5cm, vmargin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
    allcolors=black,
}
\usepackage{mathptmx} %Times font

\title{
    \textbf{
    TESSEROIDES CON DENSIDADES VARIABLES:
    CÁLCULO DE CAMPOS GRAVITATORIOS EN COORDENADAS ESFÉRICAS CON
    DENSIDADES VARIABLES EN PROFUNDIDAD
    }
}
\author{
    Santiago R. Soler$^{1,2}$ \vspace{0.5em} \\ 
    \textit{$^1$ Instituto Geofísico Sismológico Volponi, Universidad Nacional de San Juan} \\
    \textit{$^2$ Consejo Nacional de Investigaciones Científicas y Técnicas (CONICET)} \vspace{0.4em} \\
    email: santiago.r.soler@gmail.com
}
\date{}

@@

@@

\vspace{-2.5em}
\begin{center}
\textbf{Palabras Clave:} poner algo
\end{center}
\vspace{0.5em}

\begin{abstract}
El cálculo directo de campos gravitatorios para grandes extensiones requiere tomar en cuenta la curvatura de la Tierra, lo cual suele resolverse a través del uso de prismas esféricos llamados Tesseroides. Además, estructuras como cuencas sedimentarias presentan variaciones de densidad con la profundidad.

\end{abstract}

\section{Introducción}

Las variaciones de la densidad en la corteza con la profundidad han sido estudiadas por casi un siglo: \citet{Athy1930} obtuvo variaciones exponenciales, mientras que otras funciones fueron propuestas en los siguientes años por \citet{Maxant1980, Rao1986, Rao1993, Rao1994}.
También se ha realizado modelado directo o modelos de inversión gravimétricos con densidades variables, principalmente aplicados en cuencas y fosas \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

Estos modelos directos fueron desarrollados para cuerpos bidimensionales o tridimensionales en coordenadas cartesianas, los cuales se ajustan adecuadamente para aplicaciones de escalas locales.
Sin embargo, la disponibilidad de datos gravimétricos satelitales permite llevar a cabo estudios e interpretaciones a escalas regionales, en las cuales es necesario tomar en cuenta la curvatura de la Tierra a la hora de realizar modelados directos.

@@

\bibliographystyle{apalike-es}
\bibliography{../manuscript/bibtex/references}
